---
title: "VCF QC plots of clean (filtered) variants"
author: "Erin E. Chille"
date: "`r Sys.Date()`"
output:
  html_document:
  highlight: tango
  number_sections: no
  theme: default
  toc: yes
  toc_depth: 3
  toc_float:
    collapsed: no
    smooth_scroll: yes
---



# Setup

Setup R env. Load packages and set default image export formats, size and resolution.

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
rm(list = ls()) #clear environment

knitr::opts_chunk$set(echo = TRUE,
                      fig.height = 6, 
                      fig.width = 12, 
                      dev = c("png", "pdf"),
                      dpi = 1000)
library(tidyverse)
options(scipen = 999) #Prevent scientific notation

in_path <- c("1-Montipora_WGS_workflow/1-Montipora_SNP_calling/3_Visualize_VCF_QC/1-Input/2-Filtered_VCF_stats/")
out_path <- c("1-Montipora_WGS_workflow/1-Montipora_SNP_calling/3_Visualize_VCF_QC/3-Output/2-Filtered_VCF_stats/")
```


# Variant based statistics

```{r vars, echo=FALSE, message=FALSE}
var_qual.file <- file.path(in_path, "Ulithi23_ULFMv1.pop.allsites.variant_sites.vcf.lqual.gz")
var_ldepth.file <- file.path(in_path, "Ulithi23_ULFMv1.pop.allsites.subset.vcf.ldepth.gz")
var_lmiss.file <- file.path(in_path, "Ulithi23_ULFMv1.pop.allsites.variant_sites.vcf.lmiss.gz")
var_freq.file <- file.path(in_path, "Ulithi23_ULFMv1.pop.allsites.variant_sites.vcf.frq.gz")
ind_idepth.file <- file.path(in_path, "Ulithi23_ULFMv1.pop.allsites.subset.vcf.idepth.txt")
ind_imiss.file <- file.path(in_path, "Ulithi23_ULFMv1.pop.allsites.subset.vcf.imiss.txt")
ind_het.file <- file.path(in_path, "Ulithi23_ULFMv1.pop.allsites.variant_sites.vcf.het.gz")

var_qual.min  <- 30
var_depth.min <- 10
var_depth.max <- 70
var_lmiss.min  <- 0.25
var_freq.min  <- 0.05
```


## Variant quality
```{r var_qual, echo=FALSE, message=FALSE}
var_qual <-  read.table(file = var_qual.file, 
             sep = '\t', 
             header = TRUE, 
             as.is = TRUE,
             fill = TRUE)

summary(var_qual$QUAL)

var_qual.plot <- var_qual %>%
  ggplot(aes(QUAL)) +
  geom_density(fill = "dodgerblue1", colour = "black", alpha = 0.3) +
  theme_light(); var_qual.plot

var_qual.inset.plot <- var_qual %>% 
  filter(QUAL < 100) %>% 
  ggplot(aes(QUAL)) +
  xlim(0,100) +
  geom_density(fill = "dodgerblue1", colour = "black", alpha = 0.3) +
  theme_light(); var_qual.inset.plot

pdf(file.path(out_path, "Ulithi23_ULFMv1.pop.allsites.variant_sites.vcf.lqual.pdf"), height = 20, width = 20)
var_qual.plot
dev.off()

pdf(file.path(out_path, "Ulithi23_ULFMv1.pop.allsites.variant_sites.vcf.inset.lqual.pdf"), height = 20, width = 20)
var_qual.inset.plot
dev.off()
```

## Variant mean depth
```{r var_depth, echo=FALSE, message=FALSE}
var_ldepth <-  read.table(file = var_ldepth.file, 
             sep = '\t', 
             header = TRUE, 
             as.is = TRUE,
             fill = TRUE)

summary(var_ldepth$MEAN_DEPTH)

var_ldepth %>% filter(MEAN_DEPTH > 60) %>% nrow()


var_ldepth.plot <- var_ldepth %>% 
  ggplot(aes(MEAN_DEPTH)) +
  geom_density(fill = "dodgerblue1", colour = "black", alpha = 0.3) +
  theme_light() +
  geom_vline(xintercept = var_depth.min, color = "red", size=1.0) +
  geom_vline(xintercept = var_depth.max, color = "red", size=1.0); var_ldepth.plot

pdf(file.path(out_path, "Ulithi23_ULFMv1.pop.allsites.variant_sites.vcf.ldepth.pdf"), height = 20, width = 20)
var_ldepth.plot
dev.off()
```

## Variant missingness - VCFtools
```{r var_miss, echo=FALSE, message=FALSE}
var_lmiss <- read.table(file = var_lmiss.file, 
             sep = '\t', 
             header = TRUE, 
             as.is = TRUE,
             fill = TRUE)

summary(var_lmiss$F_MISS)

var_lmiss.plot <- var_lmiss %>% 
  ggplot(aes(F_MISS)) +
  geom_density(fill = "dodgerblue1", colour = "black", alpha = 0.3) +
  xlim(0,1) +
  theme_light() +
  geom_vline(xintercept = var_lmiss.min, color = "red", size=1.0); var_lmiss.plot

pdf(file.path(out_path, "Ulithi23_ULFMv1.pop.allsites.variant_sites.vcf.lmiss.pdf"), height = 20, width = 20)
var_lmiss.plot
dev.off()
```

## Minor allele frequency
```{r var_freq, echo=FALSE, message=FALSE}
var_freq <- read.table(file = var_freq.file, 
             sep = '\t', 
             header = FALSE, 
             as.is = TRUE,
             fill = TRUE, 
             col.names = c("CHROM", "POS", "N_ALLELES", "N_CHR", "A1", "A2"))

var_freq <- filter(var_freq, CHROM != "CHROM")

# find minor allele frequency
var_freq$MAF <- var_freq %>% 
  select(A1, A2) %>% 
  apply(1, function(z) min(z)) %>% 
  as.numeric()

summary(var_freq$MAF)

var_freq.plot <- var_freq %>% 
  ggplot(aes(MAF)) +
  geom_density(fill = "dodgerblue1", colour = "black", alpha = 0.3) +
  theme_light() +
  geom_vline(xintercept = var_freq.min, color = "red", size=1.0); var_freq.plot


pdf(file.path(out_path, "Ulithi23_ULFMv1.pop.allsites.variant_sites.vcf.frq.pdf"), height = 20, width = 20)
var_freq.plot
dev.off()
```


# Individual based statistics

## Mean depth per individual
```{r ind_depth, echo=FALSE, message=FALSE}
var_idepth <-  read.table(file = ind_idepth.file, 
             sep = '\t', 
             header = TRUE, 
             as.is = TRUE,
             fill = TRUE)

summary(var_idepth$MEAN_DEPTH)


var_idepth.plot <- var_idepth %>% 
  ggplot(aes(MEAN_DEPTH)) +
  geom_histogram(fill = "dodgerblue1", colour = "black", alpha = 0.3) +
  theme_light() +
  geom_vline(xintercept = var_depth.min, color = "red", size=1.0) +
  geom_vline(xintercept = var_depth.max, color = "red", size=1.0); var_idepth.plot

pdf(file.path(out_path, "Ulithi23_ULFMv1.pop.allsites.variant_sites.vcf.idepth.pdf"), height = 20, width = 20)
var_idepth.plot
dev.off()
```

## Proportion of missing data per individual
```{r}
var_imiss <- read.table(file = ind_imiss.file, 
             sep = '\t', 
             header = TRUE, 
             as.is = TRUE,
             fill = TRUE)

summary(var_imiss$F_MISS)

var_imiss.plot <- var_imiss %>% 
  ggplot(aes(F_MISS)) +
  geom_histogram(fill = "dodgerblue1", colour = "black", alpha = 0.3) +
  theme_light() +
  geom_vline(xintercept = var_lmiss.min, color = "red", size=1.0); var_imiss.plot

pdf(file.path(out_path, "Ulithi23_ULFMv1.pop.allsites.variant_sites.vcf.imiss.pdf"), height = 20, width = 20)
var_imiss.plot
dev.off()
```


## Heterozygosity and inbreeding coefficient per individual
```{r ind_het, echo=FALSE, message=FALSE}
ind_het <- read.table(file = ind_het.file, 
             sep = '\t', 
             header = TRUE, 
             as.is = TRUE,
             fill = TRUE)

ind_het.plot <- ind_het %>% 
  arrange(F) %>% 
  ggplot(aes(x = F)) +
  geom_histogram(fill = "dodgerblue", colour = "black", alpha = 0.3) +
  labs(x = "Inbreeding coefficient F") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)); ind_het.plot

summary(ind_het$F)

pdf(file.path(out_path, "Ulithi23_ULFMv1.pop.allsites.variant_sites.vcf.filt.het.pdf"), height = 20, width = 20)
ind_het.plot
dev.off()
```

